### 2024-09-20 Article/main.tex
minima

Functor constructions allow us to generalise basic concept already seen for categories. An important example of this fact are comma categories, a more general notion of slice categories (Definition~\ref{def:slice_cat}).

@@

In particular, every presheaf has an initial and a terminal object, because $\Set$ do (Example~\ref{ex:set_init_term}).

@@

\begin{prop}\label{prop:eq_are_mono}
    Let $e: E \rightarrow A$ be the arrow that equalizes $f, g : A \rightarrow B$ in a category $\cat C$. Then, $e$ is a monomorphism.
\end{prop}

\begin{proof}
    Suppose $X$ be an object and $x, y: X \rightarrow E$ be two morphisms in $\cat C$ such that $e \circ x = e \circ y$, and let $z = e \circ x$. Then, since $e$ is an equalizer, $f \circ e = g \circ e$, and $f \circ z = g \circ z$. But, for the universal property of equalizers, there must be exactly one $u: Z \rightarrow E$ such that $z = e \circ u$. It follow that $x = u$ and $y = u$, hence $x = y$.
\end{proof}

Of all monomorphism, an interesting subclass of them is the one that contains only the equalizers.

@@

\begin{example}\label{ex:po_in_set}
    \color{blue}
    In $\Set$, given two functions $f: A \rightarrow B$ and $g: A \rightarrow C$, the pushout of them is the set $X = (B \coprod C) /_\sim$, where $\sim$ is the least equivalence relation such that $f(a) \sim g(a)$ for each $a \in A$, with $\iota_g:B \rightarrow X$ and $\iota_f : C \rightarrow X$ as arrows, sending each element of the domain in the corresponding equivalence class in $X$. In particular, for each $a \in A$:
    \begin{align*}
        (\iota_g \circ f) (a)
                        &= \iota_g(f(a))    &&\\
                        &= [(f(a), 0)]           && \textit{Definition of $\iota_g$} \\
                        &= [(g(a), 1)]           && f(a) \sim g(a) \\
                        &= \iota_f(g(a))    && \textit{Definition of $\iota_f$} \\
                        &= (\iota_f \circ g) (a) &&
    \end{align*}
    % This is not sure!!
    When both $f$ and $g$ are monos (that is, injections), then we can construct the pushout taking at the same way we have done above, with $(f(a), 0) \sim (g(a), 1)$ when such $a$ exists and $(b, 0) \sim (c, 1)$ on each $b$ and $c$ with no preimage in $A$, with $\iota_f$ and $\iota_g$ injective.
    An easy way to see this fact is considering the following situation: let $f: A \rightarrow A \cup B$ and $g: A \rightarrow A \cup C$, with $A$ disjoint from $B$ and $C$, $f(a)= a$ and $g(a) = a$. Then the pushout is the object $A \cup B \cup C$, with the inclusion as arrows, that are also injective.
    A more general case is what happens considering functions $f: A \rightarrow B$ and $g: A \rightarrow C$ injective. Differently from the previous example, in this case is not possible to take just the union of codomains as the pushout, but rather the disjoint union of them and then identify the elements $f(a)$ with $g(a)$, as we have done above. In the category of sets and functions, we have the certainty that the pullback arrows are injective. In fact, taking the equivalence relation $\sim$, we have that $f(a) \sim f(a')$ if and only if $a = a'$ by hypothesis, and then $x \sim x'$ if and only if $x = x'$, then the pushout morphisms sends each element in an equivalence class composed only by the element itself, thus are injective.
    This is an interesting property that in other categories may do not hold, and will be recalled later.
\end{example}

@@

We now introduce the notion that generalize all the universal constructions defined above. In fact, initial objects, products, equalizers an pullback (dually, terminal objects, coproducts, coequalizers and pushouts) can be seen as the particular case of a certain type of construction, called limit.

@@

\begin{example}\label{ex:pb_as_limit}
    A pullback is the limit for the diagram below.
    \[
        \begin{tikzcd}
            & B \arrow[d, "g"] \\
            A \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]
    In fact, a cone for the diagram above is an object $P$ and three arrows $\phi:P \rightarrow A$, $\psi: P \rightarrow B$ and $h: P \rightarrow C$, but the latter is uniquely determined by the other ones ($f \circ \phi = h = g \circ \psi$).
    Thus, the following diagram is a cone:
    \[
        \begin{tikzcd}
            P \arrow[r, "{\psi}"] \arrow[d, "{\phi}" swap] & B \arrow[d, "g"] \\
            A \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]
    For $(P, \phi, \psi)$ to be a pullback, it must have the universal property. In other words, it has to be a limit.

@@

All the examples we provided about limits are still valid, in their dual form, for what regard colimits. In fact, initial objects, coequalizers, coproducts and pullbacks are particular cases of limits, and the way to see this fact is the same seen in previous examples.

@@

\begin{example}\label{ex:colm_of_sets}
    As we have done in Example~\ref{ex:lim_of_sets} we illustrate how to construct colimits in the category of sets. Given a small category $\cat I$, $ I = Ob(\cat I)$, and a diagram $D: \cat I \rightarrow \Set$, consider the equivalence relation $\sim$ defined on $\coprod_{i\in I} D(i)$ such that $d_i \sim d_{i'}$ if $d_i \in D(i), d_{i'} \in D(i')$ and there exists some $\phi \in \cat I(i, i')$ such that $D(\phi)(d_i) = d_{i'}$. Then, a colimit for $D$ is the set
    $$
        C = \big ( \coprod_{i \in I} D(i) \big ) / \sim
    $$
    with the inclusions as arrows.
\end{example}

@@

\begin{definition}\label{def:adh}[$\mathcal{M}$-adhesivity]
    Let $\cat C$ be a category and $\mathcal M \subseteq \Mono(\cat C)$ containing all isomorphisms, closed under composition and stable under pullbacks and pushouts (Definition~\ref{def:stab_under_pb_po}).
    Then $\cat C$ is \emph{$\mathcal M$-adhesive} if
    \begin{enumerate}
        \item every cospan $C \xrightarrow[]{g} D \xleftarrow[]{m} B$ with $m \in \mathcal M$ can be completed to a pullback ( such pullbacks are called $\mathcal M$-pullbacks);
        \item every span $C \xleftarrow{m} A \xrightarrow{f} B$ with $ m \in \mathcal M$ can be completed to a pushout (such pushouts are called $\mathcal M$-pushouts);
        \item pushouts along $\mathcal M$-arrows are $\mathcal M$-Van Kampen squares.
    \end{enumerate}
    We also say that $\cat C$ is \emph{adhesive} when it is $\Mono(\cat C)$-adhesive, and \emph{quasiadhesive} when it is $\Reg(\cat C)$-adhesive.
\end{definition}

@@

The idea behind this theorem is to simplify calculation to show that a certain category is adhesive for some subclass of monomorphisms, considering a functor from the category of which we want to prove adhesivity to a category we know it is adhesive, requiring that such functor has some properties.

\begin{proof}
    In order to prove $\mathcal M$-adhesivity of $\cat C$, we have to verify the condition in Definition~\ref{def:adh}.
    \begin{itemize}
        \item Let $C \xrightarrow[]{g} D \xleftarrow[]{m} B$ with $m \in \mathcal M$ be a cospan in $\cat C$.Applying $F$, we obtain $F(C) \xrightarrow[]{F(g)} F(D) \xleftarrow[]{F(m)} B$, with $F(m) \in \mathcal{N}$ by hypothesis. Then, there exists a pullback $(P_F, p_{F(B)}, p_{(F(D))})$ in $\cat D$, which is an $\mathcal N$-pullback (Example~\ref{ex:pb_as_limit}). Since $F$ creates pullbacks, hence lift them (Observation~\ref{obs:funct_creat_lim_then_lift}), there exist a pullback $(P, p_B, p_D)$ in $\cat C$.
        \item Let $C \xleftarrow{m} A \xrightarrow{f} B$ with $ m \in \mathcal M$ be a cospan in $\cat C$. Analogously to the previous point, applying the functor $F$ we obtain $F(C) \xleftarrow{F(m)} F(A) \xrightarrow{F(f)} F(B)$ with $ F(m) \in \mathcal N$, and there exists a $\mathcal N$-pushout $(q_{F(C)}, q_{F(B)}, F(Q))$ in $\cat D$. Since $F$ reflects pushouts, $(q_C, q_B, Q)$ is a $\mathcal{M}$-pushout in $\cat C$.
        \item the Van Kampen property of $\mathcal M$-pullbacks follows from the closure under pullbacks and pushouts of $\mathcal M$ and from the fact that $F$ reflects pullbacks.
    \end{itemize}

>>>>

Functor constructions allow us to generalise basic concepts already seen for categories. An important example of this fact are comma categories, a more general notion of slice categories (Definition~\ref{def:slice_cat}).

@@

In particular, every presheaf has an initial and a terminal object, because $\Set$ does (Example~\ref{ex:set_init_term}).

@@

\begin{prop}\label{prop:eq_are_mono}
    Let $e: E \rightarrow A$ be the arrow that equalizes $f, g : A \rightarrow B$ in a category $\cat C$. Then $e$ is a monomorphism.
\end{prop}

\begin{proof}
    Suppose $X$ be an object and $x, y: X \rightarrow E$ be two morphisms in $\cat C$ such that $e \circ x = e \circ y$, and let $z = e \circ x$. Then, since $e$ is an equalizer, $f \circ e = g \circ e$ and $f \circ z = g \circ z$. But, for the universal property of equalizers, there must be exactly one $u: Z \rightarrow E$ such that $z = e \circ u$. It follow that $x = u$ and $y = u$, hence $x = y$.
\end{proof}

Of all monomorphisms, an interesting subclass of them is the one that contains only the equalizers.

@@

\begin{example}\label{ex:po_in_set}
    \color{blue}
    In $\Set$, given two functions $f: A \rightarrow B$ and $g: A \rightarrow C$, the pushout of them is the set $X = (B \coprod C) /_\sim$, where $\sim$ is the least equivalence relation such that $f(a) \sim g(a)$ for each $a \in A$, with $\iota_g:B \rightarrow X$ and $\iota_f : C \rightarrow X$ as arrows, sending each element of the domain in the corresponding equivalence class in $X$. In particular, for each $a \in A$:
    \begin{align*}
        (\iota_g \circ f) (a)
                        &= \iota_g(f(a))    &&\\
                        &= [(f(a), 0)]           && \textit{Definition of $\iota_g$} \\
                        &= [(g(a), 1)]           && f(a) \sim g(a) \\
                        &= \iota_f(g(a))    && \textit{Definition of $\iota_f$} \\
                        &= (\iota_f \circ g) (a) &&
    \end{align*}
    % This is not sure!!
    When both $f$ and $g$ are monos (that is, injections), then we can construct the pushout in the same way we have done above, with $(f(a), 0) \sim (g(a), 1)$ when such $a$ exists and $(b, 0) \sim (c, 1)$ on each $b$ and $c$ with no preimage in $A$, with $\iota_f$ and $\iota_g$ injective.
    An easy way to see this fact is considering the following situation: let $f: A \rightarrow A \cup B$ and $g: A \rightarrow A \cup C$, with $A$ disjoint from $B$ and $C$, $f(a)= a$ and $g(a) = a$. Then the pushout is the object $A \cup B \cup C$, with the inclusions as arrows, that are also injective.
    A more general case is what happens considering functions $f: A \rightarrow B$ and $g: A \rightarrow C$ injective. Differently from the previous example, in this case is not possible to take just the union of codomains as the pushout, but rather the disjoint union of them and then identify the elements $f(a)$ with $g(a)$, as we have done above. In the category of sets and functions, we have the certainty that the pullback arrows are injective. In fact, taking the equivalence relation $\sim$, we have that $f(a) \sim f(a')$ if and only if $a = a'$ by hypothesis, and then $x \sim x'$ if and only if $x = x'$, then the pushout morphisms sends each element in an equivalence class composed only by the element itself, thus are injective.
    This is an interesting property that in other categories may do not hold, and will be recalled later.
\end{example}

@@

We now introduce the notion that generalize all the universal constructions defined above. In fact, initial objects, products, equalizers and pullbacks (dually, terminal objects, coproducts, coequalizers and pushouts) can be seen as the particular case of a certain type of construction, called limit.

@@

\begin{example}\label{ex:pb_as_limit}
    A pullback is the limit for the diagram below.
    \[
        \begin{tikzcd}
            & B \arrow[d, "g"] \\
            A \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]
    In fact, a cone for the diagram above is an object $P$ and three arrows $\phi:P \rightarrow A$, $\psi: P \rightarrow B$, and $h: P \rightarrow C$, but the latter is uniquely determined by the other ones ($f \circ \phi = h = g \circ \psi$).
    Thus, the following diagram is a cone:
    \[
        \begin{tikzcd}
            P \arrow[r, "{\psi}"] \arrow[d, "{\phi}" swap] & B \arrow[d, "g"] \\
            A \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]
    For $(P, \phi, \psi)$ to be a pullback, it must have the universal property. In other words, it has to be a limit.

@@

All the examples we provided about limits are still valid, in their dual form, for colimits. In fact, initial objects, coequalizers, coproducts and pullbacks are particular cases of limits, and the way to see this fact is the same seen in previous examples.

@@

\begin{example}\label{ex:colm_of_sets}
    As we have done in Example~\ref{ex:lim_of_sets}, we illustrate how to construct colimits in the category of sets. Given a small category $\cat I$, $ I = Ob(\cat I)$, and a diagram $D: \cat I \rightarrow \Set$, consider the equivalence relation $\sim$ defined on $\coprod_{i\in I} D(i)$ such that $d_i \sim d_{i'}$ if $d_i \in D(i), d_{i'} \in D(i')$ and there exists some $\phi \in \cat I(i, i')$ such that $D(\phi)(d_i) = d_{i'}$. Then, a colimit for $D$ is the set
    $$
        C = \big ( \coprod_{i \in I} D(i) \big ) / \sim
    $$
    with the inclusions as arrows.
\end{example}

@@

\begin{definition}\label{def:adh}[$\mathcal{M}$-adhesivity]
    Let $\cat C$ be a category and $\mathcal M \subseteq \Mono(\cat C)$ containing all isomorphisms, closed under composition and stable under pullbacks and pushouts (Definition~\ref{def:stab_under_pb_po}).
    Then $\cat C$ is \emph{$\mathcal M$-adhesive} if
    \begin{enumerate}
        \item every cospan $C \xrightarrow[]{g} D \xleftarrow[]{m} B$ with $m \in \mathcal M$ can be completed to a pullback (such pullbacks are called $\mathcal M$-pullbacks);
        \item every span $C \xleftarrow{m} A \xrightarrow{f} B$ with $ m \in \mathcal M$ can be completed to a pushout (such pushouts are called $\mathcal M$-pushouts);
        \item pushouts along $\mathcal M$-arrows are $\mathcal M$-Van Kampen squares.
    \end{enumerate}
    We also say that $\cat C$ is \emph{adhesive} when it is $\Mono(\cat C)$-adhesive, and \emph{quasiadhesive} when it is $\Reg(\cat C)$-adhesive.
\end{definition}

@@

The idea behind this theorem is to simplify calculations to show that a certain category is adhesive for some subclass of monomorphisms, considering a functor from the category of which we want to prove adhesivity to a category we know it is adhesive, requiring that such functor has some properties.

\begin{proof}
    In order to prove $\mathcal M$-adhesivity of $\cat C$, we have to verify the condition in Definition~\ref{def:adh}.
    \begin{itemize}
        \item Let $C \xrightarrow[]{g} D \xleftarrow[]{m} B$ with $m \in \mathcal M$ be a cospan in $\cat C$.Applying $F$, we obtain $F(C) \xrightarrow[]{F(g)} F(D) \xleftarrow[]{F(m)} B$, with $F(m) \in \mathcal{N}$ by hypothesis. Then, there exists a pullback $(P_F, p_{F(B)}, p_{(F(D))})$ in $\cat D$, which is an $\mathcal N$-pullback (Example~\ref{ex:pb_as_limit}). Since $F$ creates pullbacks, hence lifts them (Observation~\ref{obs:funct_creat_lim_then_lift}), there exists a pullback $(P, p_B, p_D)$ in $\cat C$.
        \item Let $C \xleftarrow{m} A \xrightarrow{f} B$ with $ m \in \mathcal M$ be a cospan in $\cat C$. Analogously to the previous point, applying the functor $F$ we obtain $F(C) \xleftarrow{F(m)} F(A) \xrightarrow{F(f)} F(B)$ with $ F(m) \in \mathcal N$, and there exists a $\mathcal N$-pushout $(q_{F(C)}, q_{F(B)}, F(Q))$ in $\cat D$. Since $F$ reflects pushouts, $(q_C, q_B, Q)$ is a $\mathcal{M}$-pushout in $\cat C$.
        \item the Van Kampen property of $\mathcal M$-pullbacks follows from the closure under pullbacks and pushouts of $\mathcal M$ and from the fact that $F$ reflects pullbacks.
    \end{itemize}
